---
title: 'Lab: Post-processing and statistics of dendrochronological series'
output:
  html_document:
    self_contained: no
    lib_dir: libs
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**During this exercise, we will standardize tree-ring series, develop a master-chronology, calculate descriptive statistics and produce relevant plots in R.**

We will use the package **dplR**.

> **Reference**: Bunn AG (2008). A dendrochronology program library in R (dplR). Dendrochronologia, 26(2), 115-124 (URL: https://doi.org/10.1016/j.dendro.2008.01.002).

> **Reference**: Bunn AG (2010). Statistical and visual crossdating in R using the dplR library. Dendrochronologia, 28(4), 251-258 (URL: https://doi.org/10.1016/j.dendro.2009.12.001).

## Data

For this exercise, we will use the *rwl* data of raw ring widths for Pacific silver firs (*Abies amabilis*) at Hurricane Ridge in the State of Washington, USA.
These data include 23 series covering 286 years and are directly available with the **dplR** package as test dataset.

We will start loading the required packages and the data.

```{r, warning = FALSE, message = FALSE}
library(dplR)
data(wa082)
str(wa082)
```

## 1. Standardization and descriptive statistics

(1a). Use the function *summary* of **dplR** to compute descriptive statistics of individual tree-ring series.
From the results, compute the mean sample length, the mean standard deviation, the mean tree-ring width, and the mean first order autocorrelation.

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
wa082.sum <- summary(wa082)
head(wa082.sum)
mean(wa082.sum$year)
mean(wa082.sum$stdev)
mean(wa082.sum$mean)
mean(wa082.sum$ar1)
```

(1b). Use the function *interseries.cor* of **dplR** to compute the individual series correlation against the master chronology and obtain the mean value.

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
head(interseries.cor(wa082))
mean(interseries.cor(wa082)[, 1])
```

(1c). Visualize rapidly the data with **dplR**. Plot the time spans of individual series (function *seg.plot*; segment plot) and visualize all tree-ring width series (function *spag.plot*; spaghetti plot). 

```{r, warning = FALSE, message = FALSE, echo=FALSE}
seg.plot(wa082)
spag.plot(wa082)
```

(1d). Standardize the data detrending each series with a spline that has a frequency response of 50% at a wavelength of 2/3 of each series's length. This step transforms tree-ring widths in indexes. To achieve this step, use the function *detrend* of **dplR** with the option *method="Spline"*. Finally visualize the spaghetti plot of these new data and compare this plot with the previous one obtained in *(1c)*.

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
wa082.rwi <- detrend(wa082, method="Spline")
str(wa082.rwi)
spag.plot(wa082.rwi)
```

## 2. Chronology building and descriptive statistics

(2a). Build a mean-value chronology by averaging the rows of the rwi data using a Tukey's biweight robust mean. To achieve this step, use the function *chron* of **dplR**. Explore the obtained chronology dataframe and plot the chronology (function *plot*) superimposing a 30-year smoothing spline for visualization (option *add.spline=TRUE, nyrs=30*).

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
wa082.crn <- chron(wa082.rwi)

str(wa082.crn)
tail(wa082.crn)

plot(wa082.crn, add.spline=TRUE, nyrs=30)
```

(2b). Compute standard statistics for chronologies with **dplR**. The first step is to correctly assign site, tree, and core IDs to each sample composing the chronology from the rwl data.frame (use the function *autoread.ids* of **dplR**). Then you compute standard statistics with the function *rwi.stats.running*. You should specify  site, tree, and core IDs of each sample and the window overlap (*window.overlap*) of contiguous 50 year blocks where the statistic are computed. The result is a data.frame containing the following columns:

**n =**
the average number of trees (for each year, a tree needs at least one non-NA core in order to be counted).

**n.tot =**
total number of correlations calculated as n.wt + n.bt.

**rbar.tot =**
the mean of all the correlations between different cores

**rbar.wt =**
the mean of the correlations between series from the same tree over all trees

**rbar.bt =**
the mean interseries correlation between all series from different trees

**c.eff =**
the effective number of cores (takes into account the number of within-tree correlations in each tree)

**rbar.eff =**
the effective signal calculated as rbar.bt / (rbar.wt + (1-rbar.wt) / c.eff)

**eps =**
the expressed population signal calculated using the average number of trees as
*n x rbar.eff / ((n - 1) x rbar.eff + 1)*

**snr =**
the signal to noise ratio calculated using the average number of trees as
*n x rbar.eff / (1-rbar.eff)*

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
wa082.ids <- autoread.ids(wa082)
head(wa082.ids)

wa082.ids.stats=rwi.stats.running(wa082.rwi, wa082.ids,window.overlap =49)
head(wa082.ids.stats)
```

(2c). Compute the subsample signal strength (SSS; *n[1+(N-1)rbar]/N[1+(n-1)rbar]*) using the function *sss* of **dplR**.

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
wa082.sss <- sss(wa082.rwi, wa082.ids)
head(wa082.sss)
```

(2d). Plot the chronology showing a potential cutoff year based on SSS or on EPS

```{r, warning = FALSE, message = FALSE}
# Plot cutoff
yrs <- time(wa082.crn)
op <- par(no.readonly=TRUE)
par(mar = c(2, 2, 2, 2), mgp = c(1.1, 0.1, 0), tcl = 0.25,mfcol = c(2, 1), xaxs='i')
plot(yrs, wa082.crn[, 1], type = "n", xlab = "Year",ylab = "RWI", axes=FALSE)
cutoff <- max(yrs[wa082.sss < 0.85])
xx <- c(500, 500, cutoff, cutoff)
yy <- c(-1, 3, 3, -1)
polygon(xx, yy, col = "grey80")
abline(h = 1, lwd = 1.5)
lines(yrs, wa082.crn[, 1], col = "grey50")
lines(yrs, ffcsaps(wa082.crn[, 1], nyrs = 30), col = "red", lwd = 2)
axis(1); axis(2); axis(3);
par(new = TRUE)
## Add SSS
plot(yrs, wa082.sss, type = "l", xlab = "", ylab = "",axes = FALSE, col = "blue",ylim=c(0,1))
abline(h=0.85,col="blue",lty="dashed")
axis(4)
mtext("SSS", side = 4, line = 1.1)
box()

# Plot cutoff2
plot(yrs, wa082.crn[, 1], type = "n", xlab = "Year",ylab = "RWI", axes=FALSE)
cutoff <- min(wa082.ids.stats$mid.year [wa082.ids.stats$eps> 0.85])-1
xx <- c(500, 500, cutoff, cutoff)
yy <- c(-1, 3, 3, -1)
polygon(xx, yy, col = "grey80")
abline(h = 1, lwd = 1.5)
lines(yrs, wa082.crn[, 1], col = "grey50")
lines(yrs, ffcsaps(wa082.crn[, 1], nyrs = 30), col = "red", lwd = 2)
axis(1); axis(2); axis(3);
par(new = TRUE)
## Add EPS
plot(wa082.ids.stats$mid.year, wa082.ids.stats$eps, type = "l", xlab = "", ylab = "",axes = FALSE, col = "blue",xlim=range(yrs),ylim=c(0,1))
abline(h=0.85,col="blue",lty="dashed")
axis(4)
mtext("EPS", side = 4, line = 1.1)
box()
par(new = TRUE)
plot(wa082.ids.stats$mid.year, wa082.ids.stats$rbar.eff, type = "l", xlab = "", ylab = "",axes = FALSE, col = 1,xlim=range(yrs),ylim=c(0,1))
```


