---
title: 'Lab: Response functions and transfer functions'
output:
  html_document:
    self_contained: no
    lib_dir: libs
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**During this exercise, we will recover climate data and we will practice with response and transfer functions in R.**

We will download climate data from **Climate Explorer**

We will use the R package **treeclim**.

## 1. Download climate data from Climate Explorer

(1a). Go to Climate Explorer

> **Weblink**: https://climexp.knmi.nl/start.cgi

(1b). Select Monthly Observations

(1c). Select CRU TS4.04 Mean Temperature : Land / 1901-2019: CRU TS 4.04 (land)

(1d). Select the study region : ave_region:: lon= -73.000 -69.000, lat= 54.000 55.000

(1e). Dowload raw data

## 2. Response functions with the package **treeclim**

We will start loading the required packages and the data, including the climate data we just extracted and a black spruce ring width chronology in the taiga of Quebec. We will use this chronology for our exercise.
We also assure to have perfect consistency for the years of these training datasets.

The ring width chronology can be found here: [Pivot.SC.crn.csv](../data/Pivot.SC.crn.csv)

```{r, warning = FALSE, message = FALSE}
library(treeclim)

# Temperature data
Cl=read.delim("../data/icru4_tmp_-73--69E_54-55N_n.dat",
              skip = 24,header=F,
              sep="")
names(Cl)=c("Year","jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec")
row.names(Cl)=Cl$Year
str(Cl)

# TR data
Pivot.SC.crn=read.csv("../data/Pivot.SC.crn.csv",row.names = 1)
str(Pivot.SC.crn)

# Assure year consistency
Pivot.SC.crn=Pivot.SC.crn[row.names(Pivot.SC.crn) %in%  Cl$Year,]
Cl=Cl[Cl$Year %in% row.names(Pivot.SC.crn),]
```

(2a). Use the function *dcc* of **treeclim** to perform, first, climate-growth correlation analysis (method = "correlation"), second, climate-growth response function analysis (method = "response"). The results can be visualized with the functions *summary*, *plot* and *barplot*. Comment the results and figures.

```{r, warning = FALSE, message = FALSE, echo=FALSE, results='hide'}
Pivot.SC.crn.cor=dcc(Pivot.SC.crn,Cl,method = "correlation",dynamic = "static")
summary(Pivot.SC.crn.cor)
plot(Pivot.SC.crn.cor)
barplot(Pivot.SC.crn.cor$coef$coef,col = Pivot.SC.crn.cor$coef$significant,names.arg = Pivot.SC.crn.cor$coef$month,las=2)

Pivot.SC.crn.res=dcc(Pivot.SC.crn,Cl,method = "response",dynamic = "static")
summary(Pivot.SC.crn.res)
plot(Pivot.SC.crn.res)
barplot(Pivot.SC.crn.res$coef$coef,col = Pivot.SC.crn.res$coef$significant,names.arg = Pivot.SC.crn.res$coef$month,las=2)
```

(2b). Use the function *dlm* of **treeclim** to verify the adjustment of the response function obtained in step 2a. In step 2a, we see that the retained months of the response function analysis are previous September, previous October, and current May, July and August. We can thus set *selection=c(-9,-10,5,7,8)* as an argument of the function. A summary is obtained with the function *summary*. What is the variance explained by the model?

```{r, warning = FALSE, message = FALSE, echo=FALSE}
Mylm2=dlm(Pivot.SC.crn,Cl,selection=c(-9,-10,5,7,8))
summary(Mylm2)
```

(2c). Produce a scatterplot of observed tree-ring data (*YOUR_RESPONSE_LIST\$truncated\$tree*) versus predicted tree ring data by the response function model (*predict(YOUR_RESPONSE_MODEL)*). Comment on the plot.

```{r, warning = FALSE, message = FALSE, echo=FALSE}
# cor(Pivot.SC.crn.res$truncated$tree,predict(Mylm2))
par(pty="s")
plot(Pivot.SC.crn.res$truncated$tree,predict(Mylm2),
     xlim=c(0.6,1.9),
     ylim=c(0.6,1.9),
     xlab="OBS",
     ylab="RESPONSE FUNCTION MODEL")
lines(c(-100,100),c(-100,100))
```

## 3. Transfer function analysis: Linear scaling #####

While a response function finds the optimal combination of climate variables to model tree-ring data, a transfer function predict a climate variable from tree ring data and may be used for a climate reconstruction. In this exercise we want to predict July-August mean temperature values from our tree ring data.

```{r, warning = FALSE, message = FALSE}
jul_aug=(Cl$jul+Cl$aug)/2
MyYear=Cl$Year
```

A dendroclimatic reconstruction has always a training calibration period that it is used to define the transfer function. The transfer function is then applied over all years with tree-ring data to reconstruct the climate (even when no meteorological instrumental data are available). Let's set a calibration period.

```{r, warning = FALSE, message = FALSE}
Calibration=1930:2016
```

The linear scaling is the simplest transfer function method that transform tree-ring data to have same mean and SD of the climate data over the calibration period with the following formula:

$REC_{i} = (TR_{i} - MeanTR_{cal}) * (SDClim_{cal}/SDTR_{cal}) + MeanClim_{cal}$

(3a). Recunstruct the 1901-2016 July-August mean temperature values based on the selected calibration period and on the linear scaling formula. Then, plot the results as shown here below (the black line is the tree-ring based reconstruction and the hatched area is the calibration period).

```{r, warning = FALSE, message = FALSE, echo=FALSE}
MeanTR=mean(Pivot.SC.crn$Pivot.SC.crn[MyYear %in% Calibration])
SDTR=sd(Pivot.SC.crn$Pivot.SC.crn[MyYear %in% Calibration])
MeanCl=mean(jul_aug[MyYear %in% Calibration])
SDCl=sd(jul_aug[MyYear %in% Calibration])

Reconstruction=(Pivot.SC.crn$Pivot.SC.crn-MeanTR)/SDTR*SDCl+MeanCl

plot(MyYear,Reconstruction,type="l",ylim=c(9,15),xlab="Year",ylab="Jul-Aug temperature")
lines(MyYear,jul_aug,col=2)
xx <- c(min(Calibration), min(Calibration), max(Calibration), max(Calibration))
yy <- c(-100, 100, 100, -100)
polygon(xx, yy, col = "grey80",density=40)
```

(3b) Compute the following statistics that may be used to assess the quality of a reconstruction:

- Correlation between the climate and the reconstruction over the full, the calibration and the verification period

- The reduction of error

  $RE = 1 - \frac{\sum_{i = 1}^{n_{ver}}{(Clim_i-Rec_i)^2}}{\sum_{i = 1}^{n_{ver}}{(Clim_i-\bar{Clim_{cal}})^2}}$

- The coefficient of efficiency

  $CE = 1 - \frac{\sum_{i = 1}^{n_{ver}}{(Clim_i-Rec_i)^2}}{\sum_{i = 1}^{n_{ver}}{(Clim_i-\bar{Clim_{ver}})^2}}$

```{r, warning = FALSE, message = FALSE, echo=FALSE}
#correlation
r1=cor(jul_aug,Reconstruction)
paste("r full period =",r1)

r2=cor(jul_aug[MyYear %in% Calibration],Reconstruction[MyYear %in% Calibration])
paste("r calibration period =",r2)

r3=cor(jul_aug[!MyYear %in% Calibration],Reconstruction[!MyYear %in% Calibration])
paste("r verification period =",r3)

#Reduction of error
RE=1-(sum((jul_aug[!MyYear %in% Calibration]-Reconstruction[!MyYear %in% Calibration])^2))/
      (sum((jul_aug[!MyYear %in% Calibration]-mean(jul_aug[MyYear %in% Calibration]))^2))
paste("RE =",RE)

#Coefficient of efficiency
CE=1-(sum((jul_aug[!MyYear %in% Calibration]-Reconstruction[!MyYear %in% Calibration])^2))/
      (sum((jul_aug[!MyYear %in% Calibration]-mean(jul_aug[!MyYear %in% Calibration]))^2))
paste("CE =",CE)

```


