---
title: 'Lab: Dendroecology at the intra annual scale: Dendrometers (solutions)'
output:
  html_document:
    self_contained: no
    lib_dir: libs
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**During this exercise, we will read, plot and start to analyze dendrometer data.**

We will use the package **fpp3**, **ggplot2**, **cowplot**.

## Data

We will start loading the required packages and the data.

```{r, warning = FALSE, message = FALSE}
library(fpp3)
library(ggplot2)
library(cowplot)
dendro=read.delim("../dendrometer/Amos_sable_dendrometres_2.dat",skip = 3,sep = ",")
dendro_names=read.delim("../dendrometer/Amos_sable_dendrometres_2.dat",skip = 1,nrows = 2,sep = ",")
names(dendro)=names(dendro_names)
rm(dendro_names)
head(dendro)
```

## 1. Create a temporal data frame (tsibble)

As a first step, you must transform the column containing the date by an usable date in R. Consult the following website to understand how to deal with date/time data in *R*: https://www.stat.berkeley.edu/~s133/dates.html. You can use the function *as.POSIXct*.

A temporal data frame can be created with the function *as_tsibble* of the **tsibble** package. A temporal data frame has an index variable with inherent ordering from past to present.

You can visualize the results with the function *head*, *plot(ts())* and *autoplot {ggplot2}*.

```{r, warning = FALSE, message = FALSE}
dendro$TIMESTAMP=as.POSIXct(dendro$TIMESTAMP)
dendro = as_tsibble(dendro, index = TIMESTAMP)
head(dendro)
plot(ts(dendro[,5:14]))
autoplot(dendro,SEVolt_Avg.2.)
```

## 2. Unit conversion

Transform the raw data in milliVolt to millimeters. See the "Cabling for dendrometers" document:

$mm = Vout/Vex * 11000 * 0.001$

```{r, warning = FALSE, message = FALSE}
dendro[,5:14]=dendro[,5:14]/2500*11000*0.001
```

## 3. Aggregation to daily values

A temporal aggregation can be fast obtained using *index_by() + summarise()* of the **tsibble** package. For this step you can simply use the observation of only one dendrometer: **SEVolt_Avg.2.** 

You can visualize the results with the function *head* and *autoplot {ggplot2}*.

```{r, warning = FALSE, message = FALSE}
dendro_daily <- index_by(dendro, day = as.Date(dendro$TIMESTAMP)) %>%
  summarize(SEVolt_Avg.2. = mean(SEVolt_Avg.2.))
head(dendro_daily)
autoplot(dendro_daily,vars(SEVolt_Avg.2.))
```

# 4. Plot time-series and daily cycles

Plot the time series of one dendrometer (**SEVolt_Avg.2.**) and of the internal datalogger temperature. Use *plot_grid {cowplot}* and *autoplot*

Plot the daily cycle of the same dendrometer (**SEVolt_Avg.2.**) and of the internal datalogger temperature. Use *plot_grid {cowplot}* and *gg_season*.

What do you deduce from these plots?

```{r, warning = FALSE, message = FALSE}
plot_grid(
  autoplot(dendro, SEVolt_Avg.2.),
  autoplot(dendro, PTemp_C_Avg),
  ncol = 1, align = "v")

plot_grid(gg_season(dendro, y = SEVolt_Avg.2., period="day"),
          gg_season(dendro, y = PTemp_C_Avg, period="day"),
          ncol = 1, align = "v")
```          
          
# 5. Autocorrelation and the partial autocorrelation of the data

Analyze the autocorrelation and the partial autocorrelation of the **SEVolt_Avg.2.** time series 

What do you deduce from these plots? (1 point)

```{r, warning = FALSE, message = FALSE}
plot_grid(autoplot(ACF(dendro, SEVolt_Avg.2.)), autoplot(PACF(dendro, SEVolt_Avg.2.)))
```  

